---
title: "Prepare data for HMM analysis"
author: "Joern Alexander Quent"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


# Preparation
## Libs and parameters
```{r load_libs}
# Libraries
library(ggplot2)
library(assortedRFunctions)
library(cowplot) 
library(stringr)
library(plyr)
library(viridis)
library(knitr)
library(ggforce)
library(gghalves)
library(ggrain)
```


<details>
 <summary>Click here for detailed session information. </summary>
```{r session_info}
sessioninfo::session_info()
```
</details>

<details>
 <summary>Click here for chunk for statistical reporting parameters. </summary>
```{r colour_and_report_params}
# Parameters how to report means
report_type   <- 1
digits1       <- 2
rounding_type <- "signif"

# Colours used for visualisation
baseColours          <- c("#003f5c", "#bc5090", "#ffa600")
```
</details>


## Loading and preparing the data
<details>
 <summary>Click here for loading data. </summary>
```{r load_data}
# Load the data
## Specify paths where the data is saved
path2data <- "data/"

## Load the look-up table that contains information of R-numbers which are retracted 
lookupTable  <- read.csv(paste0(path2data, "lookUpTable.csv"))

## Load .RData images of the combined data (all subject in one DF)
load(paste0(path2data, "demographics.RData"))
load(paste0(path2data, "DW_all_data.RData"))
load(paste0(path2data, "OLM_7T_all_data.RData"))
load(paste0(path2data, "OLM_3T_all_data.RData"))
load(paste0(path2data, "question_data.RData"))

# Select the subjects included in this analysis
## Load the subjects that are included in this analysis
subjectFile <- readLines(paste0(path2data, "SpaNov_subject2analyse.txt"))
subjIDs_R   <- str_split(subjectFile, pattern = ",")[[1]] 
subjIDs     <- lookupTable$anonKey[lookupTable$Rnum %in% subjIDs_R]
# Important note: subjIDs_R do not have the same order as subjIDs!!!!!!!!!!!!!

## Subset to data that is being included in the analysis
OLM_7T_position_data <- OLM_7T_position_data[OLM_7T_position_data$subject %in% subjIDs, ]
demographics         <- demographics[demographics$subject %in% subjIDs, ]
DW_position_data     <- DW_position_data[DW_position_data$subject %in% subjIDs, ]
OLM_7T_logEntries    <- OLM_7T_logEntries[OLM_7T_logEntries$ppid %in% subjIDs, ]
OLM_7T_trial_results <- OLM_7T_trial_results[OLM_7T_trial_results$subject %in% subjIDs, ]
question_data        <- question_data[question_data$subject %in% subjIDs, ]

## Subset to retrieval only
OLM_7T_retrieval <- OLM_7T_trial_results[OLM_7T_trial_results$trialType == "retrieval", ]
OLM_7T_encoding  <- OLM_7T_trial_results[OLM_7T_trial_results$trialType == "encoding", ]

# Get the object locations to verify object placement in screenshots
obj_locations <- ddply(OLM_7T_trial_results, c("targets", "objectName", "targetNames"),
                       summarise, object_x_sd = sd(object_x), object_x = mean(object_x),
                       object_z_sd = sd(object_z), object_z = mean(object_z))
```

</details>

# Prepare data 
For now we decided to start modelling only encoding trials because of the simplicity. For this we're using the following measures:

- Locomotion
  - Percent rotation
  - Percent stationary
  - Percent translation
- Distances
  - to the centre
  - to the object
  - traveled (probably not worth calculating)
- Number of button presses


## Locomotion
```{r locomotion}
# Subset to encoding
OLM_7T_position_encoding <- OLM_7T_position_data[OLM_7T_position_data$trialType == "encoding", ] 

# The amount we found the values to avoid false positive
rotation_round  <- 2 # round rotation values to this decimal point

# Function to determine which state a time point belongs to
what_state <- function(rot_y, moving2){
  # Get angles 
  angle1 <- rot_y[2:length(rot_y)]
  angle2 <- rot_y[1:(length(rot_y)-1)]
  
  # Calculate the amount was rotated between the time points and then rotate
  rotated <- c(NA, round(angularDifference(angle1, angle2), rotation_round))
  
  # If rotation is zero, call it stationary, otherwise it is rotation
  tra_rot_sta <- ifelse(abs(rotated) == 0 | is.na(rotated), 
                        'stationary', 'rotation') 
  
  # Set time point to translation based the information saved by unity
  tra_rot_sta[moving2] <- 'translation'
  
  # Return
  return(tra_rot_sta)
}

# Calculate the state for each time points for each subject and each trial
OLM_7T_position_encoding <- ddply(OLM_7T_position_encoding, c("subject", "trial"), 
                  mutate, locomotion = what_state(rot_y, moving))
```

## Downsampling and average measures
```{r downsample_and_calculate_measures}
# Create X-second bins
time_window <- 2
OLM_7T_position_encoding$time_bin <- floor(OLM_7T_position_encoding$time / time_window) * time_window

# Add current object position to the data
OLM_7T_position_encoding$object_x <- NA
OLM_7T_position_encoding$object_z <- NA

## Loop through and add information for each time
for(i in 1:nrow(OLM_7T_encoding)){
  # Get index for this participant and this trial
  index <- OLM_7T_position_encoding$subject == OLM_7T_encoding$subject[i] & 
    OLM_7T_position_encoding$trial == OLM_7T_encoding$trial_num[i]
 
  # Add object location to position data frame
  OLM_7T_position_encoding$object_x[index] <- OLM_7T_encoding$object_x[i]
  OLM_7T_position_encoding$object_z[index] <- OLM_7T_encoding$object_z[i]
}

# Calculate Euclidean distance between current position and object position
x1 <- OLM_7T_position_encoding$pos_x
z1 <- OLM_7T_position_encoding$pos_z
x2 <- OLM_7T_position_encoding$object_x
z2 <- OLM_7T_position_encoding$object_z
OLM_7T_position_encoding$dist2obj <- euclideanDistance3D(x1, 1, z1, x2, 1, z2)

# Calculate distance to centre
x1 <- OLM_7T_position_encoding$pos_x
z1 <- OLM_7T_position_encoding$pos_z
x2 <- 0
z2 <- 0
OLM_7T_position_encoding$dist2cenrte <- euclideanDistance3D(x1, 1, z1, x2, 1, z2)

# Downsample using ddply
OLM_7T_HMM <- ddply(OLM_7T_position_encoding, c("subject", "trial", "trialType", "time_bin"), 
                    summarise, duration = max(time) - min(time),
                    event_start = min(time), event_end = max(time), 
                    translation = mean(locomotion == "translation"),
                    rotation = mean(locomotion == "rotation"),
                    stationary = mean(locomotion == "stationary"), 
                    dist2obj = mean(dist2obj),
                    dist2cenrte = mean(dist2cenrte))

# Make start of the trial time 0
OLM_7T_HMM <- ddply(OLM_7T_HMM, c("subject", "trial"), mutate, time = time_bin - min(time_bin))

# Calculate number of button presses
## Subset the log data frame to only button presses
button_messages <- c("Datapixx: Green!", "Datapixx: Red!", "Datapixx: Yellow!", "Datapixx: Blue!")
OLM_7T_logEntries_button <- OLM_7T_logEntries[OLM_7T_logEntries$message %in% button_messages, ]

## Loop through OLM_7T_HMM and count button presses for the time periods in OLM_7T_logEntries_button
OLM_7T_HMM$button_presses <- NA
for(i in 1:nrow(OLM_7T_HMM)){
  # Get start and time plus subject
  event_start <- OLM_7T_HMM$event_start[i]
  event_end   <- OLM_7T_HMM$event_end[i]
  subject     <- OLM_7T_HMM$subject[i]
  
  # Count the number of TRUE values
  OLM_7T_HMM$button_presses[i] <- sum(OLM_7T_logEntries_button$ppid == subject & 
        (OLM_7T_logEntries_button$timestamp >= event_start & OLM_7T_logEntries_button$timestamp < event_end))
}
```

## Visualise measures
```{r visualise_measure, fig.width = 10}
# Convert from wide to long
OLM_7T_HMM_long <- reshape2::melt(OLM_7T_HMM, 
                                id.vars = c("subject", "trial", "trialType", "time_bin", "duration"))
# Locomotion
## Subset to locomotion
locomotion_df <- OLM_7T_HMM_long[OLM_7T_HMM_long$variable %in% c("translation", "rotation", "stationary"), ]

## Create plot
loco_plot <- ggplot(locomotion_df, aes(x = variable, y = value, fill = variable)) +
	geom_rain(rain.side = 'l', point.args = list(alpha = 0.01), 
	          point.args.pos = list(position = position_jitter(width = 0.09, height = 0)),
	          boxplot.args.pos = list(width = 0.05, position = position_nudge(x = 0.15)),
	          violin.args.pos  = list(width = 0.7, position = position_nudge(x = 0.20))) +
  theme_classic() +
  scale_fill_manual(values = baseColours) +
  theme(legend.position = "none") +
  labs(title = "Locomotion states", x = "Locomotion states", y = "Percent")

# Distances
## Subset to distance to centre and object
distance_df <- OLM_7T_HMM_long[OLM_7T_HMM_long$variable %in% c("dist2obj", "dist2cenrte"), ]

## Create plot
dist_plot <- ggplot(distance_df, aes(x = variable, y = value, fill = variable)) +
	geom_rain(rain.side = 'l', point.args = list(alpha = 0.01), 
	          point.args.pos = list(position = position_jitter(width = 0.09, height = 0)),
	          boxplot.args.pos = list(width = 0.05, position = position_nudge(x = 0.15)),
	          violin.args.pos  = list(width = 0.7, position = position_nudge(x = 0.20))) +
  scale_fill_manual(values = baseColours[c(1, 3)]) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = "Distances", x = "Measure", y = "Virtual metres")

# Distances
## Subset to distance to centre and object
button_df <- OLM_7T_HMM_long[OLM_7T_HMM_long$variable %in% c("button_presses"), ]

## Create plot
button_plot <- ggplot(button_df, aes(x = variable, y = value, fill = variable)) +
	geom_rain(rain.side = 'l', point.args = list(alpha = 0.01), 
	          point.args.pos = list(position = position_jitter(width = 0.09, height = 0)),
	          boxplot.args.pos = list(width = 0.05, position = position_nudge(x = 0.15)),
	          violin.args.pos  = list(width = 0.7, position = position_nudge(x = 0.20))) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_fill_manual(values = baseColours[2]) +
  labs(title = "Button\npresses", x = "Measure", y = "Number of button presses")

# combine
plot_grid(loco_plot, dist_plot, button_plot, align = "hv",
          rel_widths = 3:1, nrow = 1)
```

# Write to disk
```{r save2csv}
write.csv(OLM_7T_HMM, file = "data/OLM_HMM_encoding.csv", row.names = FALSE)
```
